Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-07-03 docs/paper.Rmd
write health and nutrition in the q-q model
Expanded the paper to include a new subsection discussing the role of health and nutrition in the quantity–quality (Q–Q) model of fertility and child investment. Added theoretical discussion, empirical evidence, and policy implications, along with new references and citations supporting the extended framework.

>>>>

The original Q–Q models often used a single catch-all variable for child quality, typically thought of as education or “expenditure per child.” Recent work emphasizes that child quality is multi-faceted, and that parents make trade-offs along several dimensions of investment—cognitive development, health, nutrition, etc. This is particularly salient in developing countries, where basic health and nutrition are pressing concerns alongside schooling.

The theoretical question is how these dimensions interact with the quantity decision. If parents allocate a budget across, say, schooling \( q_{\text{edu}} \) and nutrition/health \( q_{\text{health}} \) for each child, then having more children forces cutbacks in both dimensions (unless parents reallocate across them). In some models, health and education are complementary: a healthier child benefits more from education, and an educated mother might raise a healthier child. This complementarity can amplify the Q–Q trade-off—investing in one dimension (health) increases the returns to investing in the other (education), so a high-quality strategy becomes more focused on fewer children.

On the other hand, if one dimension has diminishing returns more quickly than another, parents might prioritize achieving a threshold level of health for all children before adding more education, which creates a nonlinear effect on fertility. One especially important aspect of health in the Q–Q framework is child survival. The probability that a child survives to adulthood effectively multiplies the utility of having that child. Historically, high child mortality led to a strategy of “quantity for insurance,” where parents had additional births to ensure survivorship. As mortality falls due to public health improvements, parents can shift toward quality without risking childlessness \citep{kalemli2002does, kalemli2000mortality}.

\citet{kalemli2002does} developed a stochastic model in which fertility choices are made under uncertainty about child survival. Her results show that declining mortality causally reduces fertility and increases educational investment per child, in line with the Q–Q trade-off. In unified growth models, declining child mortality reinforces the demand-for-human-capital channel of the demographic transition \citep{galor2004physical}.

Beyond survival, nutritional status is a critical indicator of quality in many low-income settings. Parents often face choices about how to allocate food or medical care, especially under resource scarcity. Empirical evidence supports the Q–Q model in this context: children with many siblings often exhibit poorer nutritional outcomes. A recent study using Vietnam Young Lives data finds that an additional sibling reduces height-for-age and weight-for-age z-scores by about 0.3 standard deviations on average \citep{chen2021population}  This effect is substantial, suggesting that children in larger families tend to be shorter and lighter than their peers from smaller families, likely due to the more limited distribution of healthcare and parental attention

Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia , where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs. Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health \( H \) depends on food \( F \) and medical care \( M \), such that:

\[
H = g(F, M)
\]

and \( H \) enters either utility directly or the child’s future productivity. Parents then choose \( n \), \( F \), \( M \), and possibly schooling per child. Larger \( n \) reduces \( F \) and \( M \) per child, lowering \( H \); if parents place high value on \( H \), or if \( H \) enhances returns to education, they will opt for smaller families.

Importantly, improvements in nutrition (through income growth or public programs) can first lead to both better health and higher fertility (since healthier women can bear more children). But over time, as standards of living rise, better nutrition increases the returns to investing in fewer, healthier children. Economic historian \citet{fogel1994economic} argues that Europe’s demographic transition was partly driven by better diets: improved nutrition raised the productivity of educated workers, which made education more valuable and shifted family preferences toward quality.

These propositions are important for modern development policy. Family planning programs that encourage lower fertility without improving child health and education may have limited long-term effects. Conversely, health and nutrition interventions—such as vaccinations or food supplementation—can enhance the returns to education and induce parents to reduce fertility voluntarily. The extended Q–Q framework thus serves as a unifying tool to understand how diverse interventions—ranging from school fees to nutrition programs—shape long-run development trajectories.

### 2025-07-03 docs/paper.tex
write health and nutrition in the q-q model
Expanded the paper to include a new subsection discussing the role of health and nutrition in the quantity–quality (Q–Q) model of fertility and child investment. Added theoretical discussion, empirical evidence, and policy implications, along with new references and citations supporting the extended framework.

>>>>

\subsection{Health and Nutrition in the Q–Q Model}

The original Q--Q models often used a single catch-all variable for
child quality, typically thought of as education or ``expenditure per
child.'' Recent work emphasizes that child quality is multi-faceted, and
that parents make trade-offs along several dimensions of
investment---cognitive development, health, nutrition, etc. This is
particularly salient in developing countries, where basic health and
nutrition are pressing concerns alongside schooling.

The theoretical question is how these dimensions interact with the
quantity decision. If parents allocate a budget across, say, schooling
\(q_{\text{edu}}\) and nutrition/health \(q_{\text{health}}\) for each
child, then having more children forces cutbacks in both dimensions
(unless parents reallocate across them). In some models, health and
education are complementary: a healthier child benefits more from
education, and an educated mother might raise a healthier child. This
complementarity can amplify the Q--Q trade-off---investing in one
dimension (health) increases the returns to investing in the other
(education), so a high-quality strategy becomes more focused on fewer
children.

On the other hand, if one dimension has diminishing returns more quickly
than another, parents might prioritize achieving a threshold level of
health for all children before adding more education, which creates a
nonlinear effect on fertility. One especially important aspect of health
in the Q--Q framework is child survival. The probability that a child
survives to adulthood effectively multiplies the utility of having that
child. Historically, high child mortality led to a strategy of
``quantity for insurance,'' where parents had additional births to
ensure survivorship. As mortality falls due to public health
improvements, parents can shift toward quality without risking
childlessness \citep{kalemli2002does, kalemli2000mortality}.

\citet{kalemli2002does} developed a stochastic model in which fertility
choices are made under uncertainty about child survival. Her results
show that declining mortality causally reduces fertility and increases
educational investment per child, in line with the Q--Q trade-off. In
unified growth models, declining child mortality reinforces the
demand-for-human-capital channel of the demographic transition
\citep{galor2004physical}.

Beyond survival, nutritional status is a critical indicator of quality
in many low-income settings. Parents often face choices about how to
allocate food or medical care, especially under resource scarcity.
Empirical evidence supports the Q--Q model in this context: children
with many siblings often exhibit poorer nutritional outcomes. A recent
study using Vietnam Young Lives data finds that an additional sibling
reduces height-for-age and weight-for-age z-scores by about 0.3 standard
deviations on average \citep{chen2021population} This effect is
substantial, suggesting that children in larger families tend to be
shorter and lighter than their peers from smaller families, likely due
to the more limited distribution of healthcare and parental attention

Similar patterns emerge in sub-Saharan Africa
\citep{bishwakarma2019first} and South Asia , where first-born children
experience slower growth once younger siblings arrive---again pointing
to intrahousehold trade-offs. Theoretically, one can extend the Becker
model with a nutrition production function, where a child's health \(H\)
depends on food \(F\) and medical care \(M\), such that:

\[
H = g(F, M)
\]

and \(H\) enters either utility directly or the child's future
productivity. Parents then choose \(n\), \(F\), \(M\), and possibly
schooling per child. Larger \(n\) reduces \(F\) and \(M\) per child,
lowering \(H\); if parents place high value on \(H\), or if \(H\)
enhances returns to education, they will opt for smaller families.

Importantly, improvements in nutrition (through income growth or public
programs) can first lead to both better health and higher fertility
(since healthier women can bear more children). But over time, as
standards of living rise, better nutrition increases the returns to
investing in fewer, healthier children. Economic historian
\citet{fogel1994economic} argues that Europe's demographic transition
was partly driven by better diets: improved nutrition raised the
productivity of educated workers, which made education more valuable and
shifted family preferences toward quality.

These propositions are important for modern development policy. Family
planning programs that encourage lower fertility without improving child
health and education may have limited long-term effects. Conversely,
health and nutrition interventions---such as vaccinations or food
supplementation---can enhance the returns to education and induce
parents to reduce fertility voluntarily. The extended Q--Q framework
thus serves as a unifying tool to understand how diverse
interventions---ranging from school fees to nutrition programs---shape
long-run development trajectories.
